\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-28 honors_thesis/journal_paper/paper.tex
half way through fixing formulas SOME MAY BE BROKEN
half way through cleaning up new intro

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{V}}
\providecommand{\specpos}{\ensuremath{X}}
\providecommand{\specval}{\ensuremath{FX}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
\providecommand{\casexn}{\ensuremath{(x:\neigh)}}
\providecommand{\casexx}{\ensuremath{(x:x)}}
\providecommand{\casexl}{\ensuremath{(x:\leftind)}}
\providecommand{\casexr}{\ensuremath{(x:\rightind)}}
\providecommand{\casepn}{\ensuremath{(\pers:\neigh)}}
\providecommand{\casepl}{\ensuremath{(\pers:\leftind)}}
\providecommand{\casepr}{\ensuremath{(\pers:\rightind)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are better optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO.  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

We show that while speculative evaluation in PSO is an intriguing idea, its
usefulness is limited unless some slight modifications are made to the PSO
algorithm.  When we relax the requirement that speculative evaluation exactly
reproduce the behavior of the original PSO, we can make improvements such as
using all of the extra speculative particles, instead of throwing away all but
one of them, or speculating several iterations ahead instead of just one.
Using these algorithmic modifications, we see dramatic improvements on a number
of functions over simply adding additional particles to the swarm.

\end{abstract}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  i
%Even in desktop machines, systems with 4 or 8 cores are
%commonplace.  
In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm itself and thus is
not the focus of this paper.

This previous parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.

The purpose of this paper is to consider pso parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take on the order of hours.
Our purpose is to explore the question ``since pso is often run with
a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
For the well-know benchmark function Sphere and as shown in \fig{basic-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
The effect is easier to see if we plot the value obtained as a function of swarm size, as show in 
fig{sideways-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
Other benchmark function behave similarly. As shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
Certainly these benchmark functions do not prove that this will be the case for all functions,
but it does provide some evidence that for at least some functions,
additional particles does not seem to have much benefit after some point,
some time earlier, some time later. This issue has been commented on by others
in the form of recommended swarms sizes for various types of functions (cite standard swarm paper).

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guesses wrong, execution restarts right where it would have been anyway.  This
is called speculative execution or branch prediction.

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
explicitly on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.  This could not be done in gradient methods, for instance, where the
step size for sampling depends on the gradient and therefor the position and \emph{value}
at the current position, but it
has been done in simulated annealing~\cite{witte-tpds91}.

@@

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, if we relax the requirements of the algorithm, no
longer demanding that it strictly mimic the behavior of the original PSO, we
can introduce new speculative techniques that often out-perform PSO.  Some
modifications that we explore are picking the speculative position with the
best value instead of the position that matches the branch that PSO took;
pruning some of the speculative evaluations to increase the number of particles
in the swarm; and speculating several iterations ahead instead of just one.

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once, then presents a comparison of the speculative
algorithm with the original PSO.  In \sec{relax}, we discuss various methods of
improving the performance of speculative evaluation in PSO, all of which break
the requirement of strictly reproducing the behavior of the original algorithm.
In \sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently find the optimal solution to a
given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle $i$ are updated as follows:
\begin{align}
\nonumber
	\pvel_{i,t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) \\
\label{eq:velupdate}
			& \quad \quad + \ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
\end{align}
where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
such as the personal $\pers$ qualifier to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

The PSO algorithm just described can be parallelized by having each additional
processor add a particle or a set of particles to the swarm.
But as we have seen in the introduction, for some
functions, a more desirable use of resources is to speed up the algorithm,
performing more iterations instead of adding more particles (for a fixed number of function evaluations),
previous techniques have not addressed this issue.
In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.
We do this by refactoring the standard PSO equations as described here.

%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle $i$ as $\rightn$ and $\leftn$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purpose here
so we instead use a set of indicator variables, one each representing each
possible update to the \emph{pair} $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

@@

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all cases.

For minimization, we can define this example indicator variable representing the case
where for particle $i$ will need to update $\pbest$ with the better
position $x_{i,t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
\begin{align}
\nonumber
	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\leftn_{i,t+1}),\ofunc(\rightn_{i,t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\leftn_{i,t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\rightn_{i,t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
indicator variables for the other update cases shown in Table~\ref{tab:evals}
($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
obviously using the corresponding boolean expression.

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \leftn_{i,t+1}, \rightn_{i,t+1},
        \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
\label{eq:defvocasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t+1}\otimes(\pbest_{i,t+1} - \ppos_{i,t}+1) +
			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
		\right]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
The update equation is taken exactly from the statdard update equations~\eqref{eq:velupdate}
and \eqref{eq:posupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
$\specvel_{i,t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & = \\
\label{eq:defvcasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t+1}\otimes(\ppos_{i,t+1} - \ppos_{i,t+1}) +
			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
		\right]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\nonumber
	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
\end{align}

Lastly, we also define notation for the value of the objective function at these points,
once again using $\casexn$ as an example:
\begin{align}
\label{eq:deffcasexn}
	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\leftn_{i,t+1},\rightn_{i,t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\nonumber
	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
\end{align}

With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
\begin{align}
\label{eq:vel2update}
	\pvel_{i,t+1} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t+1}^{c} \\
\label{eq:pos2update}
	\ppos_{i,t+1} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
\end{align}

>>>>

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(x,\neigh)}}
\providecommand{\casexx}{\ensuremath{(x,x)}}
\providecommand{\casexl}{\ensuremath{(x,\leftind)}}
\providecommand{\casexr}{\ensuremath{(x,\rightind)}}
\providecommand{\casepn}{\ensuremath{(\pers,\neigh)}}
\providecommand{\casepl}{\ensuremath{(\pers,\leftind)}}
\providecommand{\casepr}{\ensuremath{(\pers,\rightind)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, in many cases this approach does not scale to 
very large populations sizes (on very large clusters).
In response to this issue, we take inspiration from 
speculative execution commonly performed in modern processors and propose
Speculative Evaluation in PSO (SEPSO).

In SEPSO future positions of the particles are
speculated and evaluated in parallel with the evaluation of current positions.
We show that the set of speculative evaluations will always include the position and corresponding
evaluation that would normally have occurred at the next iteration.
Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.
Not only can SEPSO can achieve better fitness in the same number of iterations,
it also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

While Speculative Evaluation in PSO is an intriguing idea, its
usefulness is limited unless some slight modifications are made to the PSO
algorithm.  By relaxing the requirement that speculative evaluation exactly
reproduce the behavior of the original PSO, we can make better use of the information
obtained by the speculative evaluations.
Using these algorithmic modifications, we see dramatic improvements in the
effectiveness of the algorithm. 
\end{abstract}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.
%Even in desktop machines, systems with 4 or 8 cores are
%commonplace.  
In order to efficiently utilize these resources for
computationally intensive problems, PSO needs to run in parallel.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{mcnabb-cec07,belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
and asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm itself and thus is
not the focus of this paper.

These previous parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases
the number of processors never exceeds the number of particles.
The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.

The purpose of this paper is to consider pso parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least minutes, but probably hours or perhaps days.
Our purpose is to explore the question ``Since pso is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
as a function of swarm size
for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
Increasing the swarm size from 10 to 50 has a significant effect value obtained
however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
reducing the progress the algorithm makes per-evaluation.
For this reason, research recommends the use a swarm size of 50 for this function
~\cite{bratton-sis07}.
Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guesses wrong, execution restarts where it would have been anyway.
This is called speculative execution or branch prediction.

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.
This is trivially true for optimization by random sampling, for example.
The next random sample does not depend on the previous point nor on the value
of the objective function at that point in any way.
It depends only on a random number.
There is no need to wait to see
what objective function value was obtained at the previous point,
evaluation of the object function at the next point can begin as soon as the next random number is drawn.
In other optimization algorithms, the next position depends on the previous position, but not the
value obtained there, a random walk for example.
In this case
the next position to be evaluated
depends only on the previous position and a random number.
In this case, as it was in the case of random sampling,
the object function evaluation of the next position can be done in parallel
because it 
does not depend on the \emph{value} of the objective function at the previous position.
There are an infinite number of possible next positions, but only one given the draw of the random number.

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the previous position. In such
cases all of the next possible positions can be evaluated at the same time as the
evaluation of the current position.
One of these speculative positions will exactly preserve the behavior of the algorithm
and the others are wasted.
This approach has been applied to 
simulated annealing where the next position depends on (1) the last position, (2) a random number
and (3) the result of a binary test comparing the value of the objective function at the current state with the value of the objective function at the
last position.
In this context the next position can be computed in the case where the binary test is true
and in the case where it is false.
in simulated annealing to compute (1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
The test can be evaluated once all three of these evaluations of the objective function are complete, thus
computing all of the objective function evaluations
needed for two iterations in parallel and all at once~\cite{witte-tpds91}.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
By way of contrast, this approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and \emph{value}
at the current position. In this case there are an infinite number of possible speculative
evaluations.

@@

In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
a speculative evaluation approach like the ones described above. We prove that this approach
produces exactly equivalent results, and demonstrate this by implementation.
The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
surprisingly, with fewer total function evaluations.

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
We can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce pso we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard pso would have done.

The balance of this paper is organized as follows:
\sec{pso} describes the particle swarm optimization algorithm.
\sec{sepso}
describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once, then presents a comparison of the speculative
algorithm with the original PSO.  In \sec{relax}, we discuss various methods of
improving the performance of speculative evaluation in PSO, all of which break
the requirement of strictly reproducing the behavior of the original algorithm.
In \sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently approximate the optimal solution to a
given objective function.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal $\pers$ qualifier to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing needed for each
particle on a separate processor.
But as we have seen in the introduction, for some
functions, and for large numbers of processors where just adding particles (processors)
reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness appreciably faster.

To address this issue we present here a very different use of additional processors.
Performing more iterations instead of adding more particles (for a fixed number of function evaluations),
previous techniques have not addressed this issue.
In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.
We do this by refactoring the standard PSO equations as described here.

%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purposes
so we instead use a set of indicator variables, one each representing each
possible update to the \emph{pair} $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

@@

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all cases.

For minimization, we can define this example indicator variable representing the case
where the particle will need to update $\pbest$ with the better
position $x_{t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
\begin{align}
\nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{t}^{\casexn}$.  The
indicator variables for the other update cases shown in Table~\ref{tab:evals}
($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
obviously using the corresponding boolean expression.

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvocasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
The update equation is taken exactly from the standard update equations~\eqref{eq:velupdate}
and \eqref{eq:posupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t+1}$ we can re-write
$\specvel_{t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
\nonumber
	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}

Lastly, we use the normal notation for the value of the objective function at these points,
once again using the case $\casexn$ as an example:
\begin{align}
\nonumber
	\ofunc(\specpos_{t+1}^{\casexn} & (\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t}))= \\
\label{eq:deffcasexn}
	& =  \ofunc(\specpos_{t+1}^{\casexn})
\end{align}
